AHABus \& FCORE: A Modular Bus System for High-Altitude Balloon Missions

### 2017-04-18 chapters/03-methodology.tex
First draft

In this chapter, the \acrfull{ahabus} is presented as a prototype solution to the issues exposed in chapter \ref{ch:lit-review}.

@@

The proposed solution for this project is inspired by the bus model that
supports the architecture of most commercial satellites. A standard software
and hardware platform provides common facilities required on all High-Altitude
flights – position tracking, internal data transfers and radio telemetry to the
ground station – allowing payloads (scientific instruments for example) to
only generate data and make it available to the platform. A global overview of the platform and the environment used for its development is presented, followed by a detailed discussion of the development process of each of the four main components of \acrshort{ahabus} (Radio Communications, Payload Bus, Flight Software and Ground Segment). Finally, the methods used to evaluate the viability of the prototype platform are presented, leading to the results presented in chapter \ref{ch:results}.

@@

This section describes the design process that led to the radio communications
protocol used by the AHABus platform to send payload data and flight metadata
to the ground station. Its role is to ensure that arbitrary binary data – each
payload can use their own data format – as stably and reliably as possible. As
discussed in the following subsections, the design and implementation choice
were also driven by legal constraints: in the United Kingdom, airborne radio
transmitters are limited in frequency band (70cm wavelength) and power
(transmission power inferior to 10mW).

@@

\subsection{AHABus Radio Protocol}

The AHABus radio protocol was designed to match the requirements for the radio
system as closely as possible. While the specification presented in this section
is the last version, which was implemented in the AHABus prototype, some parts
of the protocol evolved or had to be changed during the prototyping process.
% TODO: Might not be necessary to say that here, given further sections

@@

Other existing protocols exist that can transmit data over radio audio signals,
like DominoEx or Olivia, sometimes with better transmission rates. RTTY was
preferred over them because the ESP8266 board used as OBC in the AHABus
prototype provides hardware support for UART, making it easier to implement and
reliable. Since the link layer is transparent to the frame and packet layers,
RTTY could be swapped for another protocol in the future with no impact on
those.

@@

Identifying lost sections of an arbitrary stream of bytes without any form of
formatting is not possible: since it is not known in advance what data should
be expected, there is no model to which actual received data can be compared.
For this reason, the AHABus radio protocol was designed as a framed protocol:
the stream of data is divided in chunks of a set number of bytes, encapsulated
in \textit{frames} that contain the a sequence number used to track the position
of the chunk in the data stream.

The AHABus frame design was inspired by other frame-based protocols like the
CCSDS Space Packet/Frame protocol. Each frame is composed of two section: a
header, that contains metadata, and a payload containing the data carried by
that frame.

@@

\begin{figure}[H]
    \begin{center}
    \begin{bytefield}[bitwidth=0.5em]{64}
        \bitheader{0,8,24,40} \\
        \bitbox{8}{version} &
        \bitbox{16}{seq. number} &
        \bitbox{40}{data[253]}
    \end{bytefield}
    \end{center}
    \centering
    \caption{AHABus Frame Format – Version 1}
    \label{fig:frame-fmt-orig}
\end{figure}

@@

\begin{figure}[H]
    \begin{center}
    \begin{bytefield}[bitwidth=0.5em]{64}
        \bitheader{0,8,16,24,40} \\
        \bitbox{8}{\texttt{0xAA}} & \bitbox{8}{\texttt{0x5A}} & \bitbox{8}{version} &
        \bitbox{16}{seq. number} &
        \bitbox{24}{data[251]}
    \end{bytefield}
    \end{center}
    \centering
    \caption{AHABus Frame Format – Version 2}
    \label{fig:frame-fmt-2}
\end{figure}

@@

\begin{figure}[H]
    \begin{center}
    \begin{tikzpicture}[scale=0.2]
    \tikzstyle{every node}+=[inner sep=0pt]
    \draw [black] (23.9,-13.4) circle (3);
    \draw (23.9,-13.4) node {$Wait$};
    \draw [black] (53.5,-13.4) circle (3);
    \draw (53.5,-13.4) node {$Sync$};
    \draw [black] (38.6,-30.7) circle (3);
    \draw (38.6,-30.7) node {$Frame$};
    \draw [black] (26.719,-12.375) arc (107.79277:72.20723:39.209);
    \fill [black] (50.68,-12.37) -- (50.07,-11.65) -- (49.77,-12.61);
    \draw (38.7,-10) node [above] {$\texttt{byte=0xAA}$};
    \draw [black] (53.11,-16.371) arc (-12.09889:-69.37588:18.591);
    \fill [black] (41.48,-29.87) -- (42.41,-30.06) -- (42.05,-29.12);
    \draw (49.56,-26.05) node [right] {$\texttt{byte=0x5A}$};
    \draw [black] (35.918,-29.359) arc (-119.94439:-159.34576:25.489);
    \fill [black] (24.79,-16.26) -- (24.61,-17.19) -- (25.54,-16.84);
    \draw (28.67,-25.22) node [left] {$\texttt{read=256}$};
    \draw [black] (51.972,-15.968) arc (-3.02212:-291.02212:2.25);
    \draw (47.22,-18.39) node [left] {$\texttt{byte=0xAA}$};
    \fill [black] (50.53,-13.75) -- (49.76,-13.21) -- (49.71,-14.21);
    \draw [black] (25.724,-11.023) arc (137.60109:42.39891:17.571);
    \fill [black] (25.72,-11.02) -- (26.63,-10.77) -- (25.89,-10.1);
    \draw (38.7,-4.8) node [above] {$\texttt{byte!=0xAA}$};
    \end{tikzpicture}
    \end{center}
    \centering
    \caption{AHABus Frame Synchronisation State Machine}
    \label{fig:frame-fsm}
\end{figure}

Figure \ref{fig:frame-fsm} presents the finite state machine that
AHABus frame decoders must implement to detect frames in a binary
stream, and figure \ref{fig:frame-fmt-3} the final version of the
AHABus frame format.

@@

\begin{figure}[H]
    \begin{bytefield}{16}
        \bitheader{0,7,8,15} \\
        \begin{leftwordgroup}{Header}
            \bitbox{8}{\texttt{0x5A}} & \bitbox{8}{version} \\
            \bitbox{16}{sequence number}
        \end{leftwordgroup} \\
        \begin{leftwordgroup}{Payload}
            \wordbox{2}{data[]}
        \end{leftwordgroup} \\
        \begin{leftwordgroup}{FEC}
            \wordbox{2}{checksum[32]}
        \end{leftwordgroup}
    \end{bytefield}
    \centering
    \caption{AHABus Frame Format - Version 3}
    \label{fig:frame-fmt-3}
\end{figure}

@@

To achieve those goals, the AHABus protocol uses packets similar to frames. Each
packet encapsulates a chunk of data retrieved from a single payload. To that
end, packets were designed to be of variable length, to avoid unnecessary
padding, which would have wasted bandwidth. The data - referred to as the
packet's \textit{payload} – is preceded by the packet's header, which contains
information necessary to attribute the packet's data to a payload, indicate
the length of the packet, and carry the AHABus platform's location and altitude.

@@

\begin{figure}[H]
    \begin{bytefield}{32}
        \bitheader{0,7,8,15,16,23,24,31} \\
        \begin{leftwordgroup}{Header}
            \bitbox{8}{version} & \bitbox{8}{payload ID} &
            \bitbox{16}{length} \\
            \bitbox{32}{latitude} \\
            \bitbox{32}{longitude} \\
            \bitbox{16}{altitude} & \bitbox[lrt]{16}{}
        \end{leftwordgroup} \\
        \begin{leftwordgroup}{Payload}
            \wordbox[lrb]{3}{data[length]}
        \end{leftwordgroup}
    \end{bytefield}
    \centering
    \caption{AHABus Packet Format - Version 1}
    \label{fig:packet-fmt-1}
\end{figure}

Figure \ref{fig:packet-fmt-1} shows the first and current version of the AHABus
packet format. Like frame discussed in section \ref{sssec:frame-layer}, each
packet carries a version flag which was added to allow backwards compatibility,
were the protocol to evolve in the future.

@@

\item In the case where a frame is lost due to channel noise or interferences,
finding the start of the next packet would have been complicated, or required
a synchronisation mechanism similar to the one designed for AHABus frames.
Aligning packets instead makes such a system unnecessary, reducing the packet
header size and the complexity of packet decoders.
\end{itemize}

@@

To evaluate the protocol's viability, a full demonstration of an AHABus radio
stack was written. Its main goals were to provide a reference implementation of
a compliant encoder and decoder, validate the protocol's design process, and
test the algorithms used in a controlled environment before integrated testing.

The C language (ISO/IEC 9899:1999 standard) was chosen for the reference
implementation because of its widespread use, and its availability on most
micro-controller and embedded platforms, which would be used as AHABus flight
platforms. To render the code as portable and reusable as possible, care was
taken to limit the dependencies to a subset of the C language standard library,
and avoid use of dynamic memory allocation, which is often inefficient or
unavailable on embedded platforms.

@@

\subsection{AHABus Payload Bus Protocol}

@@

The AHABus Payload Bus protocol was designed to work with a virtual register
system: each payload can be thought of like a memory array, and the bus
controller (the \acrshort{obc}) writes and reads data to and from addresses on
each payload. Addressing a specific payload on the bus, replying to the bus
controller and read/write operations are handled by the underlying
\acrshort{i2c} protocol (\cite{NXPSemiconductors2014}). Table
\ref{tab:bus-protocol-registers} gives the addresses, sizes and purposes of
each register. Multi-byte values are always encoded in big-endian order.

@@

\end{tabular}
\end{center}
\caption {AHABus Payload Bus Protocol - Virtual Registers}
\label{tab:bus-protocol-registers}
\end{table}

@@

\begin{figure}[H]
\includegraphics[width=0.4\textwidth]{ahabus-schema.png}
\centering
\caption{AHABus - Electronic Circuit}
\label{fig:ahabus-schema}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=0.4\textwidth]{ahabus-pcb.png}
\centering
\caption{AHABus - Printed Circuit Board Layout}
\label{fig:ahabus-pcb}
\end{figure}

@@

In order to allow a decoding tool to work on the full binary stream, a fork of
Dl-Fldigi was modified so that the binary stream could be extracted. When the
modified version starts, it creates a local network socket, to which any
byte decoded from RTTY streams are forwarded. This design was chosen because
most programming languages provide support for network communications, which
means that tools could easily be written to work with AHABus frame streams, such
as the packet parser discussed in the following section.

@@

@@

For all tests, the experimental setup was kept consistent: an ESP8266
development board was used as the \acrshort{obc}, while two Arduino Uno boards
were connected as payloads, generating fake instrument data. A \SI{5}{\volt}
power supply was connected to the power lines.

>>>>

In this chapter, the \acrfull{ahabus} is presented as a prototype solution to the issues exposed in chapter \ref{ch:literature-review}. A global overview of the platform and the environment used for its development is presented, followed by a detailed discussion of the development process of each of the four main components of \acrshort{ahabus} (Radio Communications, Payload Bus, Flight Software and Ground Segment). Finally, the methods used to evaluate the viability of the prototype platform are presented, leading to the results presented in chapter \ref{ch:results}.

@@

The proposed solution for this project is inspired by the bus model that
supports the architecture of most commercial satellites. A standard software
and hardware platform provides common facilities required on all High-Altitude
flights – position tracking, internal data transfers and radio telemetry to
the ground station – allowing payloads (scientific instruments for example) to
only generate data and make it available to the platform.

@@

This section describes the design process that led to the radio communications
protocol used by the \acrshort{ahabus} platform to send payload data and flight metadata
to the ground station. Its role is to ensure that arbitrary binary data – each
payload can use their own data format – as stably and reliably as possible. As
discussed in the following subsections, the design and implementation choice
were also driven by legal constraints: in the United Kingdom, airborne radio
transmitters are limited in frequency band (70cm wavelength) and power
(transmission power inferior to 10mW).

@@

\subsection{\acrshort{ahabus} Radio Protocol}

The \acrshort{ahabus} radio protocol was designed to match the requirements for the radio
system as closely as possible. While the specification presented in this section
is the last version, which was implemented in the \acrshort{ahabus} prototype, some parts
of the protocol evolved or had to be changed during the prototyping process.
% TODO: Might not be necessary to say that here, given further sections

@@

Other existing protocols exist that can transmit data over radio audio signals,
like DominoEx or Olivia, sometimes with better transmission rates. RTTY was
preferred over them because the ESP8266 board used as OBC in the \acrshort{ahabus}
prototype provides hardware support for UART, making it easier to implement and
reliable. Since the link layer is transparent to the frame and packet layers,
RTTY could be swapped for another protocol in the future with no impact on
those.

@@

Identifying lost sections of an arbitrary stream of bytes without any form of
formatting is not possible: since it is not known in advance what data should
be expected, there is no model to which actual received data can be compared.
For this reason, the \acrshort{ahabus} radio protocol was designed as a framed protocol:
the stream of data is divided in chunks of a set number of bytes, encapsulated
in \textit{frames} that contain the a sequence number used to track the position
of the chunk in the data stream.

The \acrshort{ahabus} frame design was inspired by other frame-based protocols like the
CCSDS Space Packet/Frame protocol. Each frame is composed of two section: a
header, that contains metadata, and a payload containing the data carried by
that frame.

@@

\begin{figure}[H]
    \begin{center}
    \begin{bytefield}[bitwidth=0.5em]{64}
        \bitheader{0,8,24,40} \\
        \bitbox{8}{version} &
        \bitbox{16}{seq. number} &
        \bitbox{40}{data[253]}
    \end{bytefield}
    \end{center}
    \centering
    \caption{\acrshort{ahabus} Frame Format – Version 1}
    \label{fig:frame-fmt-orig}
\end{figure}

@@

\begin{figure}[H]
    \begin{center}
    \begin{bytefield}[bitwidth=0.5em]{64}
        \bitheader{0,8,16,24,40} \\
        \bitbox{8}{\texttt{0xAA}} & \bitbox{8}{\texttt{0x5A}} & \bitbox{8}{version} &
        \bitbox{16}{seq. number} &
        \bitbox{24}{data[251]}
    \end{bytefield}
    \end{center}
    \centering
    \caption{\acrshort{ahabus} Frame Format – Version 2}
    \label{fig:frame-fmt-2}
\end{figure}

@@

\begin{figure}[H]
    \begin{center}
    \begin{tikzpicture}[scale=0.2]
    \tikzstyle{every node}+=[inner sep=0pt]
    \draw [black] (23.9,-13.4) circle (3);
    \draw (23.9,-13.4) node {$Wait$};
    \draw [black] (53.5,-13.4) circle (3);
    \draw (53.5,-13.4) node {$Sync$};
    \draw [black] (38.6,-30.7) circle (3);
    \draw (38.6,-30.7) node {$Frame$};
    \draw [black] (26.719,-12.375) arc (107.79277:72.20723:39.209);
    \fill [black] (50.68,-12.37) -- (50.07,-11.65) -- (49.77,-12.61);
    \draw (38.7,-10) node [above] {$\texttt{byte=0xAA}$};
    \draw [black] (53.11,-16.371) arc (-12.09889:-69.37588:18.591);
    \fill [black] (41.48,-29.87) -- (42.41,-30.06) -- (42.05,-29.12);
    \draw (49.56,-26.05) node [right] {$\texttt{byte=0x5A}$};
    \draw [black] (35.918,-29.359) arc (-119.94439:-159.34576:25.489);
    \fill [black] (24.79,-16.26) -- (24.61,-17.19) -- (25.54,-16.84);
    \draw (28.67,-25.22) node [left] {$\texttt{read=256}$};
    \draw [black] (51.972,-15.968) arc (-3.02212:-291.02212:2.25);
    \draw (47.22,-18.39) node [left] {$\texttt{byte=0xAA}$};
    \fill [black] (50.53,-13.75) -- (49.76,-13.21) -- (49.71,-14.21);
    \draw [black] (25.724,-11.023) arc (137.60109:42.39891:17.571);
    \fill [black] (25.72,-11.02) -- (26.63,-10.77) -- (25.89,-10.1);
    \draw (38.7,-4.8) node [above] {$\texttt{byte!=0xAA}$};
    \end{tikzpicture}
    \end{center}
    \centering
    \caption{\acrshort{ahabus} Frame Synchronisation State Machine}
    \label{fig:frame-fsm}
\end{figure}

Figure \ref{fig:frame-fsm} presents the finite state machine that
\acrshort{ahabus} frame decoders must implement to detect frames in a binary
stream, and figure \ref{fig:frame-fmt-3} the final version of the
\acrshort{ahabus} frame format.

@@

\begin{figure}[H]
    \begin{bytefield}{16}
        \bitheader{0,7,8,15} \\
        \begin{leftwordgroup}{Header}
            \bitbox{8}{\texttt{0x5A}} & \bitbox{8}{version} \\
            \bitbox{16}{sequence number}
        \end{leftwordgroup} \\
        \begin{leftwordgroup}{Payload}
            \wordbox{2}{data[]}
        \end{leftwordgroup} \\
        \begin{leftwordgroup}{FEC}
            \wordbox{2}{checksum[32]}
        \end{leftwordgroup}
    \end{bytefield}
    \centering
    \caption{\acrshort{ahabus} Frame Format - Version 3}
    \label{fig:frame-fmt-3}
\end{figure}

@@

To achieve those goals, the \acrshort{ahabus} protocol uses packets similar to frames. Each
packet encapsulates a chunk of data retrieved from a single payload. To that
end, packets were designed to be of variable length, to avoid unnecessary
padding, which would have wasted bandwidth. The data - referred to as the
packet's \textit{payload} – is preceded by the packet's header, which contains
information necessary to attribute the packet's data to a payload, indicate
the length of the packet, and carry the \acrshort{ahabus} platform's location and altitude.

@@

\begin{figure}[H]
    \begin{bytefield}{32}
        \bitheader{0,7,8,15,16,23,24,31} \\
        \begin{leftwordgroup}{Header}
            \bitbox{8}{version} & \bitbox{8}{payload ID} &
            \bitbox{16}{length} \\
            \bitbox{32}{latitude} \\
            \bitbox{32}{longitude} \\
            \bitbox{16}{altitude} & \bitbox[lrt]{16}{}
        \end{leftwordgroup} \\
        \begin{leftwordgroup}{Payload}
            \wordbox[lrb]{3}{data[length]}
        \end{leftwordgroup}
    \end{bytefield}
    \centering
    \caption{\acrshort{ahabus} Packet Format - Version 1}
    \label{fig:packet-fmt-1}
\end{figure}

Figure \ref{fig:packet-fmt-1} shows the first and current version of the \acrshort{ahabus}
packet format. Like frame discussed in section \ref{sssec:frame-layer}, each
packet carries a version flag which was added to allow backwards compatibility,
were the protocol to evolve in the future.

@@

\item In the case where a frame is lost due to channel noise or interferences,
finding the start of the next packet would have been complicated, or required
a synchronisation mechanism similar to the one designed for \acrshort{ahabus} frames.
Aligning packets instead makes such a system unnecessary, reducing the packet
header size and the complexity of packet decoders.
\end{itemize}

@@

To evaluate the protocol's viability, a full demonstration of an \acrshort{ahabus} radio
stack was written. Its main goals were to provide a reference implementation of
a compliant encoder and decoder, validate the protocol's design process, and
test the algorithms used in a controlled environment before integrated testing.

The C language (ISO/IEC 9899:1999 standard) was chosen for the reference
implementation because of its widespread use, and its availability on most
micro-controller and embedded platforms, which would be used as \acrshort{ahabus} flight
platforms. To render the code as portable and reusable as possible, care was
taken to limit the dependencies to a subset of the C language standard library,
and avoid use of dynamic memory allocation, which is often inefficient or
unavailable on embedded platforms.

@@

\subsection{\acrshort{ahabus} Payload Bus Protocol}

@@

The \acrshort{ahabus} Payload Bus protocol was designed to work with a virtual register
system: each payload can be thought of like a memory array, and the bus
controller (the \acrshort{obc}) writes and reads data to and from addresses on
each payload. Addressing a specific payload on the bus, replying to the bus
controller and read/write operations are handled by the underlying
\acrshort{i2c} protocol (\cite{NXPSemiconductors2014}). Table
\ref{tab:bus-protocol-registers} gives the addresses, sizes and purposes of
each register. Multi-byte values are always encoded in big-endian order.

@@

\end{tabular}
\end{center}
\caption {\acrshort{ahabus} Payload Bus Protocol - Virtual Registers}
\label{tab:bus-protocol-registers}
\end{table}

@@

\begin{figure}[H]
\includegraphics[width=0.4\textwidth]{ahabus-schema.png}
\centering
\caption{\acrshort{ahabus} - Electronic Circuit}
\label{fig:ahabus-schema}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=0.4\textwidth]{ahabus-pcb.png}
\centering
\caption{\acrshort{ahabus} - Printed Circuit Board Layout}
\label{fig:ahabus-pcb}
\end{figure}

@@

In order to allow a decoding tool to work on the full binary stream, a fork of
Dl-Fldigi was modified so that the binary stream could be extracted. When the
modified version starts, it creates a local network socket, to which any
byte decoded from RTTY streams are forwarded. This design was chosen because
most programming languages provide support for network communications, which
means that tools could easily be written to work with \acrshort{ahabus} frame streams, such
as the packet parser discussed in the following section.

@@

\begin{figure}[H]
\includegraphics[width=0.40\textwidth]{dl-fldigi.png}
\centering
\caption{\acrshort{ahabus} - Ground Segment Software}
\end{figure}

@@

For all tests, the experimental setup, as shown in figure \ref{fig:exp-setup} was kept consistent: an ESP8266 development board was used as the \acrshort{obc}, while two Arduino Uno boards were connected as payloads, generating fake instrument data. A \SI{5}{\volt} power supply was connected to the power lines.

\begin{figure}[H]
\includegraphics[width=0.40\textwidth]{test-rig.jpg}
\centering
\caption{\acrshort{ahabus} - Test Setup}
\label{fig:exp-setup}
\end{figure}
